Finite-size scaling of free energy in the dimer model on a hexagonal domain

### 2018-11-26 AffineLieAlgebras/MQFT2018/paper.tex
Submitted to TMPh

\newcommand{\Li}{\mathrm{Li}}
\newcommand{\dx}{\mathrm{dx}~}
\newcommand{\dy}{\mathrm{dy}~}
\newcommand{\dz}{\mathrm{dz}}
\newcommand{\go}{\stackrel{\circ }{\mathfrak{g}}}
\newcommand{\ao}{\stackrel{\circ }{\mathfrak{a}}}
\newcommand{\co}[1]{\stackrel{\circ }{#1}}
\newcommand{\pia}{\pi_{\mathfrak{a}}}
\newcommand{\piab}{\pi_{\mathfrak{a}_{\bot}}}
\newcommand{\gf}{\mathfrak{g}}
\newcommand{\nf}{\mathfrak{g}}
\newcommand{\gfh}{\hat{\mathfrak{g}}}
\newcommand{\af}{\mathfrak{a}}
\newcommand{\afh}{\hat{\mathfrak{a}}}
\newcommand{\bff}{\mathfrak{b}}
\newcommand{\slnn}{\mathfrak{sl}_{n}}
\newcommand{\sln}[1]{\mathfrak{sl}_{#1}}
\newcommand{\afb}{\mathfrak{a}_{\bot}}
\newcommand{\hf}{\mathfrak{h}}
\newcommand{\hfg}{\hf_{\gf}}
\newcommand{\hfa}{\hf_{\af}}
\newcommand{\hfb}{\mathfrak{h}_{\bot}}
\newcommand{\pf}{\mathfrak{p}}
\newcommand{\aft}{\widetilde{\mathfrak{a}}}
\newcommand{\sfr}{\mathfrak{s}}
%\newcommand{\dim}{\mathrm{dim} }

\begin{document}
\title{Finite-size scaling of free energy in dimer model on hexagonal domain}

\author{A.A.~Nazarov$^{1}$\\
{\small
  $^{1}$Department of Physics, St. Petersburg State University,} \\
{\small  Ulyanovskaya 1, 198504 St.~Petersburg, Russia}\\
%\small{$^{2}$}\\
%\small{$^{3}$}\\
\small{email:antonnaz@gmail.com}
}
\date{}
\maketitle

\begin{abstract}
    We consider dimer model on a hexagonal lattice. This model can be seen as a ``pile of cubes in the
  corner''. The partition function is given by the determinant of Kasteleyn matrix. We use the
  expression for the partition function to derive the scaling behavior of free energy in the limit
  of lattice mesh tending to zero and temperature tending to infinity. We discuss the connection of
  the expansion coefficients to the geometry of the domain.
\end{abstract}

@@

We consider a particular case of the dimer model on a hexagonal domain of hexagonal lattice, that
can be seen as a ``pile of cubes in the corner''. The energy of configuration is the total number of
cubes. For this particular case we use an exact combinatorial formula for the partition function to
derive the expressions for scaling limit of free energy and first three terms of the finite-size
corrections. We show that the first term is identically zero. The second term has a universal part
that is given by the Euler constant and a geometry-dependent part that is written explicitly in
elementary functions. The third term which encapsulates logarithmic dependence on the mesh size is
also shown to be universal.

We than compare these results with our previous numeric simulations \cite{belov2018finite}. 
\section{Model definition}
\label{sec:model-definition}
The configurations of the dimer model are perfect matchings (sets of non-touching edges, covering
all the vertices) on some graph ${\cal G}$ with some choice of weights $\omega(e)$ on the edges. The
model is solvable on the bipartite graphs, i.e. the partition function can be computed if the
weights are introduced in such a way that each face with 0 mod 4 edges there is an odd number of
negative edge weights and each face with 2 mod 4 edges has an even number of of negative edge
weights. Then the signs of the edge weights form a so-called ``Kasteleyn orientation''on graph, the
weighting is called ``Kasteleyn weighting'' \cite{kenyon2001dominos,kenyon2009lectures}.

For a bipartite graph ${\cal G}$, color the vertices black and white in such a way that all the
vertices adjacent to the black one are white. Denote by $W$,$B$ the sets of white and black
vertices and by $w,b$ the elements of these sets.

@@

Kasteleyn matrix defines a discrete Dirac operator $D$, in vertex $v$ we have:
\begin{equation}
  \label{eq:16}
  (Df)(v)=\sum_{u}{\cal K}(v,u) f(u)
\end{equation}

@@

MacMahon formula is obtained for the following definition of the Kasteleyn matrix. Embed the
hexagonal lattice in the complex plane $\mathbb{C}$ in such a way that some edges are parallel to
the real line. Then take
\begin{eqnarray}
  \label{eq:18}
  {\cal K}(w,b)=q^{\Re w+\Im w} \quad \mathrm{if}\quad \Im w=\Im b\\
  {\cal K}(w,b)=1 \quad \mathrm{if}\quad \Im w\neq\Im b
\end{eqnarray}

%% The scaling limit is achieved on the infinite lattice as $T\to\infty$. To study the scaling behavior
%% using Monte-Carlo simulations on the finite lattices one can consider temperatures proportional to
%% the lattice size and, then, extrapolate them to infinity.
%%   
%  \frametitle{Kasteleyn matrix and Dirac operator}
%%  Consider infinite-dimensional graded algebra, e.g. $\gf[[t]]$ or $\hat\gf$

%  \frametitle{Free energy scaling}
The   free energy per site is defined as$^{*}$
\footnote{For convenience we omit the factor $\frac{1}{T}$ in the usual definition of the free energy.}
\begin{equation*}
  \label{eq:17}
  % \frac{f}{T}
  f=-\frac{1}{V}\ln Z(M,N,K,q)
\end{equation*}
Here $V$ is the number of vertices, it is twice the number of dimers and twice the number of  cube faces:
\begin{equation*}
  \label{eq:19}
  V=2(MN+NK+MK)
\end{equation*}

%!!!! Kenyon uses $V=\#Dimers$.

We are interested in the scaling limit, combined with the thermodynamic limit, when
$ T\to \infty, \quad M,N,K\to \infty$, such that ratios
$\frac{M}{T}=a,\quad \frac{N}{T}=b, \quad \frac{K}{T}=c$ remain fixed. In what follows we use
$\varepsilon=\frac{1}{T}$, which can be seen as the scale of the model, e.g. mesh size.

@@

%%series and split the
%%logarithm of product into the sum of logarithms. Thus we get:
%%\begin{multline}
%%  \label{eq:2}
%%   -\frac{f}{T}= \sum_{i=1}^{M} \sum_{j=1}^{N} \sum_{k=1}^{K} \frac{1}{V}\left\{
%%    \ln\left(1-\frac{1}{e^{\varepsilon(i+j+k)}-1}\sum_{l=1}^{\infty} \frac{\varepsilon^{l}}{l!}\right)-\right.\\
%%  -\left.\ln\left(1-\frac{1}{e^{\varepsilon(i+j+k)}-1}\sum_{l=1}^{\infty} \frac{2^{l}\varepsilon^{l}}{l!}\right)\right\}
%%\end{multline}
%%Then expand the logarithms up to $\varepsilon^{3}$:
\begin{multline}
  \label{eq:4}
  %\frac{f}{T}
  f=- \sum_{i=1}^{M} \sum_{j=1}^{N} \sum_{k=1}^{K} \frac{1}{V}\left\{
    g(i\varepsilon,j\varepsilon,k\varepsilon)+
    \frac{3}{2}\varepsilon\left[g(i\varepsilon,j\varepsilon,k\varepsilon)+g(i\varepsilon,j\varepsilon,k\varepsilon)^{2}\right]\right.\\
  \left.+\frac{7}{6}\varepsilon^{2}\left[g(i\varepsilon,j\varepsilon,k\varepsilon)+3 g(i\varepsilon,j\varepsilon,k\varepsilon)^{2}+
      2 g(i\varepsilon,j\varepsilon,k\varepsilon)^{3}\right]\right\}  + \mathcal{O}(\varepsilon^{3})
\end{multline}

Here and below we will use the notation:
%Denote by $g$ the expression that will appear multiple times:
\begin{equation}
  \label{eq:5}
  g(x,y,z)=\frac{1}{e^{x+y+z}-1}
\end{equation}

@@

%%But we need to obtain
%%corrections in $\varepsilon$ to the integrals as $\varepsilon\to 0$.
%%
%%We can use the following approximation formula for integral of some function $\tilde{f}$ to take
%%care of the corrections:

@@

%Now we can apply the integral representation.

@@

%%Substituting this approximation and using integral representation for the quadratic
%%  term in $\varepsilon$, we obtain for the free energy:
%%  \begin{multline*}
%%    \frac{f}{T}=-\frac{1}{ab+bc+ca}\left\{\int_{0}^{a} \int_{0}^{b}\int_{0}^{c}\frac{\dx \dy \dz}{e^{x+y+z}-1}+\right.\\
%%    \left.+\frac{\varepsilon^{2}}{12}\int_{0}^{a} \int_{0}^{b}\int_{0}^{c}
%%      \left(\frac{e^{2(x+y+z)}+e^{x+y+z}}{\left(e^{x+y+z}-1\right)^{3}}-\frac{1}{bc}\frac{1}{x}\right)\dx \dy \dz+\right.\\
%%    \left.+\frac{\varepsilon^{2}}{12}(\ln a-\gamma)+\frac{\varepsilon^{2}}{12}\ln \varepsilon\right\}    
%%  \end{multline*}
%%The integral can be taken explicitly, it is equal to
%%\begin{multline}
%%  \label{eq:7}
%%  \int_{0}^{a} \int_{0}^{b}\int_{0}^{c}
%%  \left(\frac{e^{2(x+y+z)}+e^{x+y+z}}{\left(e^{x+y+z}-1\right)^{3}}-\frac{1}{bc}\frac{1}{x}\right)\dx~\dy~\dz=\\
%%  =\ln \left(\frac{(e^{a}-1)(e^{b}-1)(e^{c}-1)(e^{a+b+c}-1)}{a (e^{a+b}-1)(e^{b+c}-1)(e^{a+c}-1)}\right)=\\
%%  =\ln \left(\frac{(e^{a}-1)(e^{b}-1)(e^{c}-1)(e^{a+b+c}-1)}{(e^{a+b}-1)(e^{b+c}-1)(e^{a+c}-1)}\right) -\ln a.
%%\end{multline}

@@

The expansion  (\ref{eq:3}) can be written as
\begin{equation}
  \label{eq:26}
  f=f_{0}+f_{1}\varepsilon + f_{2}\varepsilon^{2} +f_{3}\varepsilon^{2}\ln\varepsilon,
\end{equation}
where the coefficients are
\begin{equation}
  \label{eq:27}
  f_{0}=-\frac{1}{2(ab+bc+ca)}\int_{0}^{a} \int_{0}^{b}\int_{0}^{c}\frac{\dx \dy \dz}{e^{x+y+z}-1},
\end{equation}
\begin{equation}
  \label{eq:28}
  f_{1}=0,
\end{equation}
\begin{equation}
  \label{eq:29}
  f_{2}=-\frac{1}{2(ab+bc+ca)}\frac{1}{12}\left[\ln\left(\frac{(e^{a}-1)(e^{b}-1)(e^{c}-1)(e^{a+b+c}-1)}{(e^{a+b}-1)(e^{b+c}-1)(e^{a+c}-1)}\right)-
    \gamma\right],
\end{equation}
and
\begin{equation}
  \label{eq:30}
  f_{3}=-\frac{1}{2(ab+bc+ca)}\frac{1}{12}
\end{equation}
In the scaling limit $\varepsilon\to 0$ so called ``limit shape phenomenon''~\cite{kenyon2006dimers}
appears in the dimer model. The areas around the corners of the domain are ``frozen'' with height
function values being fixed. An analytical ``Arctic curve'' delimits frozen regions from the region
where the behavior is described by the effective free field theory
\cite{kenyon2009lectures,kenyon2008height,kenyon2006dimers}.

@@

First two terms $f_{0}$ and $f_{1}$ are interpreted as a bulk and
boundary free energies in the corresponding field theory. Since we have $f_{1}=0$, we can conclude
that boundary tension is zero.
The value of the first term $f_{0}$ can be rewritten using polylogarithm functions as
\begin{multline}
  \label{eq:34}
  f_{0}=\frac{1}{2(ab+bc+ca)}\left[abc + \Li_{3}(e^{a})+\Li_{3}(e^{b})+\Li_{3}(e^{c})-
    \Li_{3}(e^{a+b})\right.\\
  \left.-\Li_{3}(e^{b+c})-    \Li_{3}(e^{a+c})+    \Li_{3}(e^{a+b+c})-\zeta(3)\right]
\end{multline}

The third term $f_{2}$ depends only on the shape of the domain through $a,b,c$ with a universal
contribution that is equal to the Euler constant $\gamma$. We conjecture that Euler constant will
appear even for a different choice of Kasteleyn weighting. In the future work we will consider
coordinate dependent values of $q$, as was done in the paper \cite{okounkov2007random}.

@@

%% In the table we present the results of computations of free energy for the case $a=b=c=1$ and
%% various values of $\varepsilon$ using an exact formula and expansion (\ref{eq:3}). In the third column
%% difference is presented. We see that the difference decays faster than $\varepsilon^{3}$. The expansion
%% (\ref{eq:3}) has good precision for comparison with Monte-Carlo simulation results such as presented
%% in
%% \cite{belov2018finite}. 
%% 
%% \begin{table}[htpb]
%% \begin{center}
%% \begin{tabular}{c|ccc}
%%   $\varepsilon$&Exact value of $f$&$f$ from expansion (\ref{eq:3})&Difference\\
%%   \hline
%%   0.2 & 0.06734815 & 0.06559065 & 0.00175751\\
%%   0.1 & 0.06689788 & 0.06681834 & 0.00007954\\
%%   0.05 & 0.06721738 & 0.06719747 & 0.00001992\\
%%   0.02 & 0.06732867 & 0.06732549 & $3.18791\cdot 10^{-6}$\\
%%   0.01 & 0.06734815 & 0.06734736 & $7.97026 \cdot 10^{-7}$
%%   0.005 & 0.06735375 & 0.06735355 & $1.99259 \cdot 10^{-7}$
%% \end{tabular}
%% \caption{Values of the free energy for $a=b=c=1$ computed by the exact formula (\ref{eq:20}) and an
%%   approximate formula (\ref{eq:3}). The difference decays faster than $\varepsilon^{3}$.
%% }
%% \label{tttt1}
%% \end{center}
%% \end{table}
%% 
%%

%%   \begin{figure}[h!tb]
%% 
%%     \includegraphics{exact-vs-expansion.pdf}
%%     \caption{Plot of the expression \eqref{eq:3} for $\varepsilon=\frac{1}{15}\dots\frac{1}{2}$, $a=3,b=2,c=1$
%%       exact values. Note that difference between exact and approximate values decreases with
%%       $\varepsilon$.}
%%   \end{figure}
%% 
%%\section{Scaling in the non-uniform model}
%%\label{sec:scaling-non-uniform}
%%
%%Consider a skew 3D partition with the partition function
%%\begin{equation}
%%  \label{eq:25}
%%  Z({q_{t}},\lambda)=\sum_{\lambda(t)}\prod_{t\in\mathbb{Z}}q_{t}^{|\lambda(t)|}
%%\end{equation}
%%
\section*{Conclusion and outlook}
\label{sec:conclusion}

@@

In further work we will show the connection of the expansion coefficients with the spectral
properties of Dirac operator on the non-frozen domain and study the universality of the presented
expressions by considering the model with the non-uniform Kasteleyn weighting with $q$ depending on
the coordinate.

@@

This research is supported by RFBR grant No. 18-01-00916.
%The calculations were carried out using the facilities of the SPbU Resource Center ``Computational Center of SPbU''.

>>>>

\newcommand{\Li}{\mathrm{Li}}
\newcommand{\dx}{\mathrm{dx}~}
\newcommand{\dy}{\mathrm{dy}~}
\newcommand{\dz}{\mathrm{dz}}

\begin{document}
\title{Finite-size scaling of free energy in the dimer model on a hexagonal domain}

\author{A.A.~Nazarov$^{1}$\\
{\small
  $^{1}$Department of Physics, St. Petersburg State University,} \\
{\small  Ulyanovskaya 1, 198504 St.~Petersburg, Russia}\\
\small{email:antonnaz@gmail.com}
}
\date{}
\maketitle

\begin{abstract}
  We consider dimer model on a hexagonal lattice. This model can be seen as a ``pile of cubes in the
  corner''. The energy of configuration is given by the volume of the pile and the partition
  function is computed by the classical MacMahon formula or, more formally, by the determinant of
  Kasteleyn matrix. We use the expression for the partition function to derive the scaling behavior
  of free energy in the limit of lattice mesh tending to zero and temperature tending to infinity.
  We discuss the universality and physical meaning of expansion coefficients.
\end{abstract}

@@

We consider a particular case of the dimer model on a hexagonal domain of hexagonal lattice, that
can be seen as a ``pile of cubes in the corner''. The energy of configuration is the total number of
cubes. For this particular case we use an exact combinatorial formula for the partition function to
derive the expressions for scaling limit of free energy and first three terms of the finite-size
corrections. We show that the first term is identically zero. The second term has a universal part
that is given by the Euler constant and a geometry-dependent part that is written explicitly in
terms of elementary functions. The third term which encapsulates logarithmic dependence on the mesh
size is connected with the central charge of the effective field theory.

This result is supported by numeric simulations presented in our previous publication
\cite{belov2018finite}.

\section{Model definition}
\label{sec:model-definition}
The configurations of the dimer model are perfect matchings (sets of non-touching edges, covering
all the vertices) on some graph ${\cal G}$ with some choice of weights $\omega(e)$ on the edges. The
model is solvable on the bipartite graphs, i.e. the partition function can be computed if the
weights are introduced in such a way that for each face bounded by 0 mod 4 edges there is an odd
number of negative edge weights and each face with 2 mod 4 edges has an even number of of negative
edge weights. Then the signs of the edge weights form a so-called ``Kasteleyn orientation''on graph,
the weighting is called ``Kasteleyn weighting'' \cite{kenyon2001dominos,kenyon2009lectures}.

For a bipartite graph ${\cal G}$, color the vertices black and white in such a way that all the
vertices adjacent to the black one are white. Denote by $B, W$ the sets of black and white
vertices and by $b,w$ the elements of these sets.

@@

Kasteleyn matrix defines a discrete Dirac operator $D$, the action of $D$ on a function $f$ defined
on vertices is given by:
\begin{equation}
  \label{eq:16}
  (Df)(v)=\sum_{u}{\cal K}(v,u) f(u)
\end{equation}

@@

MacMahon formula is obtained for the following definition of the Kasteleyn matrix. Embed the
hexagonal lattice in the complex plane $\mathbb{C}$ in such a way that some edges are parallel to
the real line with and corresponding vertices have coordinates with integer real and imaginary
parts. Then take
\begin{eqnarray}
  \label{eq:18}
  {\cal K}(w,b)=q^{\Re w+\Im w} \quad \mathrm{if}\quad \Im w=\Im b\\
  {\cal K}(w,b)=1 \quad \mathrm{if}\quad \Im w\neq\Im b
\end{eqnarray}

The   free energy per site is defined as$^{*}$
\footnote{For convenience we omit the factor $\frac{1}{T}$ in the usual definition of the free energy.}
\begin{equation*}
  \label{eq:17}
  % \frac{f}{T}
  f=-\frac{1}{V}\ln Z(M,N,K,q)
\end{equation*}
Here $V$ is the number of vertices, it is twice the number of dimers and twice the number of  cube faces:
\begin{equation*}
  \label{eq:19}
  V=2(MN+NK+MK)
\end{equation*}

We are interested in the scaling limit, combined with the thermodynamic limit, when $ T\to \infty,$
and $M,N,K\to \infty$, such that ratios $\frac{M}{T}=a,\quad \frac{N}{T}=b, \quad \frac{K}{T}=c$
remain fixed. In what follows we use $\varepsilon=\frac{1}{T}$, which can be seen as the scale of
the model, e.g. mesh size.

@@

\begin{multline}
  \label{eq:4}
  %\frac{f}{T}
  f=- \sum_{i=1}^{M} \sum_{j=1}^{N} \sum_{k=1}^{K} \frac{1}{V}\left\{
    g(i\varepsilon,j\varepsilon,k\varepsilon)+
    \frac{3}{2}\varepsilon\left[g(i\varepsilon,j\varepsilon,k\varepsilon)+g(i\varepsilon,j\varepsilon,k\varepsilon)^{2}\right]\right.\\
  \left.+\frac{7}{6}\varepsilon^{2}\left[g(i\varepsilon,j\varepsilon,k\varepsilon)+3 g(i\varepsilon,j\varepsilon,k\varepsilon)^{2}+
      2 g(i\varepsilon,j\varepsilon,k\varepsilon)^{3}\right]\right\}  + \mathcal{O}(\varepsilon^{3})
\end{multline}

Here and below we will use the notation:
\begin{equation}
  \label{eq:5}
  g(x,y,z)=\frac{1}{e^{x+y+z}-1}
\end{equation}

@@

@@

@@

@@

The expansion  (\ref{eq:3}) can be written as
\begin{equation}
  \label{eq:26}
  f=f_{0}+f_{1}\varepsilon + f_{2}\varepsilon^{2} +f_{3}\varepsilon^{2}\ln\varepsilon,
\end{equation}
where the coefficients are
\begin{equation}
  \label{eq:27}
  f_{0}=-\frac{1}{2(ab+bc+ca)}\int_{0}^{a} \int_{0}^{b}\int_{0}^{c}\frac{\dx \dy \dz}{e^{x+y+z}-1},
\end{equation}
\begin{equation}
  \label{eq:28}
  f_{1}=0,
\end{equation}
\begin{equation}
  \label{eq:29}
  f_{2}=-\frac{1}{2(ab+bc+ca)}\frac{1}{12}\left[\ln\left(\frac{(e^{a}-1)(e^{b}-1)(e^{c}-1)(e^{a+b+c}-1)}{(e^{a+b}-1)(e^{b+c}-1)(e^{a+c}-1)}\right)-
    \gamma\right],
\end{equation}
and
\begin{equation}
  \label{eq:30}
  f_{3}=-\frac{1}{2(ab+bc+ca)}\frac{1}{12}
\end{equation}
In the paper \cite{belov2018finite} we have presented results of Monte Carlo simulations using
Metropolis and Wang-Landau algorithms that support the form of expansion (\ref{eq:26}). In
particular we got $f_{1}=0.04\pm 0.04$ which is consistent with $f_{1}=0$.

In the scaling limit $\varepsilon\to 0$ so called ``limit shape phenomenon''~\cite{kenyon2006dimers}
appears in the dimer model. The areas around the corners of the domain are ``frozen'' with height
function values being fixed. An analytical ``Arctic curve'' delimits frozen regions from the region
where the behavior is described by the effective free field theory
\cite{kenyon2009lectures,kenyon2008height,kenyon2006dimers}.

@@

First two terms $f_{0}$ and $f_{1}$ are interpreted as a bulk and
boundary free energies in the corresponding field theory. Since we have $f_{1}=0$, we can conclude
that boundary tension is zero.
The value of the first term $f_{0}$ can be rewritten using polylogarithm functions as
\begin{multline}
  \label{eq:34}
  f_{0}=\frac{1}{2(ab+bc+ca)}\left[abc + \Li_{3}(e^{a})+\Li_{3}(e^{b})+\Li_{3}(e^{c})-
    \Li_{3}(e^{a+b})\right.\\
  \left.-\Li_{3}(e^{b+c})-    \Li_{3}(e^{a+c})+    \Li_{3}(e^{a+b+c})-\zeta(3)\right]
\end{multline}
Here Riemann zeta function appears as a particular value of polylogarithm $\Li_{3}(1)=\zeta(3)$.

The third term $f_{2}$ depends only on the shape of the domain through $a,b,c$ with a universal
contribution that is equal to the Euler constant $\gamma$. We conjecture that Euler constant will
appear even for a different choice of Kasteleyn weighting, due to the treatment of logarithmic
divergency. In the future work we will consider coordinate dependent values of $q$ and more complex
geometries, as was done in the paper \cite{okounkov2007random}, to check this suggestion.

@@

\section*{Conclusion and outlook}
\label{sec:conclusion}

@@

In further work we will show the connection of the expansion coefficients with the spectral
properties of Dirac operator on the non-frozen domain and study the universality of the presented
expressions by considering the model on more generic domain geometries with the non-uniform
Kasteleyn weighting with $q$ depending on the coordinate.

@@

This research is supported by RFBR grant No. 18-01-00916.
